---
title: "Preliminary geometric morphometric analysis of Gary dart points"
author: "Robert Z. Selden, Jr."
date: "December 5, 2019"
output: github_document
bibliography: gahagan.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Introduction

This document includes supplemental materials for the article, "A Quantitative Assessment of Intraspecific Morphological Variation in Gary Dart Points: An Exercise in Archaeological Epistemology and Heuristics"

### Load geomorph and data

```{r load geomorph, setwd, and load data, echo=TRUE}
# load geomorph
devtools::install_github("geomorphR/geomorph", ref = "Stable", build_vignettes = TRUE)
library(geomorph)

# set working directory
setwd(getwd())
source('readmulti.csv.R')

# read .csv files
setwd("./data")
filelist <- list.files(pattern = ".csv")
coords<-readmulti.csv(filelist)
setwd("../")

# read qualitative data
qdata<-read.csv("qdata.csv",header=TRUE, row.names=1)
qdata<-qdata[match(dimnames(coords)[[3]],rownames(qdata)),]
```

### Generalised Procrustes Analysis

```{r gpa}
Y.gpa<-gpagen(coords, PrinAxes = TRUE, print.progress = FALSE)
# plot(Y.gpa)
# quantify global integration relative to self-similarity (Bookstein 2015)
globalIntegration(Y.gpa$coords)
# geomorph data frame
gdf<-geomorph.data.frame(shape=Y.gpa$coords, size=Y.gpa$Csize, region=qdata$region)
csz<-Y.gpa$Csize # attribute for boxplot
rgn<-qdata$region # attribute for boxplot
# boxplot of central Texas and southern Caddo area Gahagan bifaces by centroid size
boxplot(csz~rgn,
        main = "Centroid size of Gary dart points by temporal assignment",
        names = c("Archaic", "Woodland"),
        xlab = "Temporal Assignment",
        ylab = "Centroid Size",
        col = c("dodgerblue4","indianred4")
        )
# plot consensus configuration in 2D to illustrate axial twisting
par(mfrow=c(1,3))
plot(Y.gpa$consensus[,c("Y", "X")], pch=20)
plot(Y.gpa$consensus[,c("Z", "X")], pch=20)
plot(Y.gpa$consensus[,c("Z", "Y")], pch=20)
# insert 3D GPA plot

```

### Principal Components Analysis

```{r pca}
# principal components analysis
PCA<-gm.prcomp(Y.gpa$coords)
summary(PCA)
# set plot parameters
colors <- c("dodgerblue4","indianred4") # baseball colors, naturally
colors <- colors[as.numeric(qdata$region)]
shapes <- c(2,4) # open triangle (ctx) and X (sca)
shapes <- shapes[as.numeric(qdata$region)]
# plotPCA
PCAplot<-plot(PCA, col = colors, pch = shapes)
```

### Define models

```{r define-models}
# define models
fit.size<-procD.lm(shape ~ size, data = gdf, print.progress = FALSE, iter = 9999)
fit.sizetime<-procD.lm(size ~ time, data = gdf, print.progress = FALSE, iter = 9999)
fit.shapetime<-procD.lm(shape ~ time, data = gdf, print.progress = FALSE, iter = 9999)
fit.unique<-procD.lm(shape ~ size * time, data = gdf, print.progress = FALSE, iter = 9999) # unique allometries
fit.common<-procD.lm(shape ~ size + time, data = gdf, print.progress = FALSE, iter = 9999) # common allometries
```

### Allometry

```{r allometry}
# allometry: does shape change with size?  
anova(fit.size)
anova(fit.common)
anova(fit.unique)
plot(fit.size, type = "regression", reg.type = "RegScore", predictor = log(gdf$size), pch = shapes, col = colors)
    # common allometric component (Mitteroecker 2004)
plotAllometry(fit.size, size = gdf$size, logsz = TRUE, method = "CAC", pch = shapes, col = colors)
    # size-shape PCA (Mitteroecker 2004)
plotAllometry(fit.shaperegion, size = gdf$size, logsz = TRUE, method = "size.shape", pch = shapes, col = colors)
    # do Gahagan biface forms from different regions express parallel, convergent, or divergent morphological characteristics?
extremes<-plotAllometry(fit.unique, size = gdf$size, logsz = TRUE, method = "PredLine", pch = shapes, col = colors)
# max/min for each population in the above result accomplished manually using:
# picknplot.shape(extremes)
anova(fit.common, fit.unique, print.progress = FALSE)
```

### Size/Shape ~ Time?

```{r region}
# ANOVA: does Gahagan biface size differ by region?
anova(fit.sizetime)
# ANOVA: do Gahagan biface shapes differ by region?
anova(fit.shapetime)
```

### Morphological disparity

```{r m-disparity}
# morphological disparity: do either of the groups display greater shape variation among individuals relative to the other group?
morphol.disparity(fit.shapetime, groups = qdata$time, data = gdf, print.progress = FALSE, iter = 9999)
# morphological disparity: do either of the groups display greater size variation among individuals relative to the other group?
morphol.disparity(fit.sizetime, groups = qdata$time, data = gdf, print.progress = FALSE, iter = 9999)
```

### Mean shapes

```{r m-shapes}
#subset landmark coordinates to produce mean shapes for groups
new.coords<-coords.subset(A = Y.gpa$coords, group = qdata$region)
names(new.coords)
#group shape means
mean<-lapply(new.coords, mshape)
# plot(mean$CTX)
# plot(mean$SCA)
# plotRefToTarget(mean$SCA,mean$CTX, method="vector",mag=2)

```

### Acknowledgments


### References cited
