---
title: "Traditional linear metrics for Gary dart points"
author: "Robert Z. Selden, Jr."
date: "February 22, 2020"
output: github_document
bibliography: landmarking-protocol.bib
fig_caption: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Introduction

This document includes supplemental materials for the article, “A quantitative assessment of intraspecific morphological variation in Gary dart points: An exercise in archaeological epistemology and heuristics.” The traditional linear measures of maximum length, width, and thickness were combined with stem length and stem width for this study. These variables represent the full suite of metrics used to characterise Gary points in previous studies, while also providing those data needed to examine whether specimens included in this study may transcend the threshold between dart and arrow points [@RN20881]. 

Others have noted a gradual diminution in the size of Gary dart points through time [@RN20898;@RN3132;@RN20874], thus it was necessary to begin by asking whether the size of those points aggregated for this study may fall below the arbitrary 11.8 mm dart-arrow index (DAI) threshold posited by Hildebrandt and King [-@RN20881]. The temporal span associated with morphologically-diagnostic Gary dart points is thought to transcend the cultural shift from hunter-gatherer to emergent horticulturalist, and overlaps temporally with the advent and flourescence of the bow and arrow. The dart-arrow index was calculated for all specimens of Gary dart points used in this study, and none fell below the dart-arrow threshold posited by Hildebrandt and King [-@RN20881]. It is noteworthy that others [@RN20880] have found that the DAI may not be universally applicable; however, the DAI and the threshold identified by Hildebrandt and King [-@RN20881] is a useful heuristic in this instance.

### Load packages for analysis

```{r load packages}
# install required analysis packages
devtools::install_github("vqv/ggbiplot")
devtools::install_github("mlcollyer/RRPP")
devtools::install_github("tidyverse/ggplot2")
devtools::install_github("kassambara/ggpubr")
# load libraries
library(ggbiplot)
library(RRPP)
library(ggplot2)
library(ggpubr)
```

### Set working directory, load data, and define variables

```{r setwd, ld, define variables, echo=TRUE}
# set working directory
setwd(getwd())
ppgary<-read.csv("garymorphlm.csv",header = TRUE)
# define variables
maxl<-ppgary$maxl # maximum length
maxw<-ppgary$maxw # maximum width
maxth<-ppgary$maxth # maximum thickness
maxstl<-ppgary$maxstl # maximum stem length
maxstw<-ppgary$maxstw # maximum stem width
dai<-ppgary$dai # dart-arrow index value
```

### Gary type-variety linear metrics

```{r gantt, echo=TRUE}
# Ford, Phillips, and Haag (1955)
fphl<-data.frame(Name=c('var.GaryStemmed-length','var.TypicalGaryStemmed-length','var.BroadGaryStemmed-length','var.ThinGaryStemmed-length','var.SmallGaryStemmed-length'),
           Length=c(45,45,45,55,35), # in mm
           end=c(70,79,70,75,50) # in mm
)
fphlength<-ggplot(fphl,aes(x=Length,xend=end,y=Name,yend=Name,color=Name)) +
  geom_segment(size=2) +
  scale_color_brewer(palette = "Dark2") +
  theme(legend.position = "none")
fphw<-data.frame(Name=c('var.GaryStemmed-width','var.TypicalGaryStemmed-width','var.BroadGaryStemmed-width','var.ThinGaryStemmed-width','var.SmallGaryStemmed-width'),
           Width=c(22,22,32,30,20), # in mm
           end=c(31,31,48,36,30) # in mm
)
fphwidth<-ggplot(fphw,aes(x=Width,xend=end,y=Name,yend=Name,color=Name)) +
  geom_segment(size=2) +
  scale_color_brewer(palette = "Dark2") +
  theme(legend.position = "none")
# render figure
fwebbfig<-ggarrange(fphlength,fphwidth,
                  labels = c("A","B"),
                  ncol = 1, nrow = 2)
annotate_figure(fwebbfig,
                top=text_grob("Gary type-variety linear metrics per Ford, Phillips, and Haag (1955)")
)
# Ford and Webb (1956)
fwebbl<-data.frame(Name=c('var.GaryLarge-length','var.GaryMed/Typical-length','var.GarySmall-length'),
           Length=c(80,46,33), # in mm
           end=c(140,79,45) # in mm
)
fwebblength<-ggplot(fwebbl,aes(x=Length,xend=end,y=Name,yend=Name,color=Name)) +
  geom_segment(size=2) +
  scale_color_brewer(palette = "Dark2") +
  theme(legend.position = "none")
fwebbw<-data.frame(Name=c('var.GaryLarge-width','var.GaryMed/Typical-width','var.GarySmall-width'),
           Width=c(30,20,19), # in mm
           end=c(42,45,32) # in mm
)
fwebbwidth<-ggplot(fwebbw,aes(x=Width,xend=end,y=Name,yend=Name,color=Name)) +
  geom_segment(size=2) +
  scale_color_brewer(palette = "Dark2") +
  theme(legend.position = "none")
# thickness values were not precisely reported for GaryLarge and GaryTypical, and are included here for reference only
fwebbth<-data.frame(Name=c('var.GaryLarge-width','var.GaryMed/Typical-width','var.GarySmall-width'),
           Thickness=c(13,9,5), # in mm
           end=c(13,10,10) # in mm
)
fwebbthickness<-ggplot(fwebbth,aes(x=Thickness,xend=end,y=Name,yend=Name,color=Name)) +
  geom_segment(size=2) +
  scale_color_brewer(palette = "Dark2") +
  theme(legend.position = "none")
# render figure
fwebbfig<-ggarrange(fwebblength,fwebbwidth,fwebbthickness,
                  labels = c("A","B","C"),
                  ncol = 1, nrow = 3)
annotate_figure(fwebbfig,
                top=text_grob("Gary type-variety linear metrics per Ford and Webb (1956)")
)

# Schambach (1998)
schambl<-data.frame(Name=c('var.Gary-length','var.Malvern-length','var.LeFlore-length','var.Bodcaw-length','var.Manice-length','var.CamdenA-length','var.CamdenB-length'),
           Length=c(51,43,43,40,36,39,50), # in mm
           end=c(73,72,80,60,57,67,80) # in mm
)
length<-ggplot(schambl,aes(x=Length,xend=end,y=Name,yend=Name,color=Name)) +
  geom_segment(size=2) +
  scale_color_brewer(palette = "Dark2") +
  theme(legend.position = "none")
schambw<-data.frame(Name=c('var.Gary-width','var.Malvern-width','var.LeFlore-width','var.Bodcaw-width','var.Manice-width','var.CamdenA-width','var.CamdenB-width'),
           Width=c(31,23,25,21,22,16,26), # in mm
           end=c(45,33,54,36,41,27,38) # in mm
)
width<-ggplot(schambw,aes(x=Width,xend=end,y=Name,yend=Name,color=Name)) +
  geom_segment(size=2) +
  scale_color_brewer(palette = "Dark2") +
  theme(legend.position = "none")
schambth<-data.frame(Name=c('var.Gary-thickness','var.Malvern-thickness','var.LeFlore-thickness','var.Bodcaw-thickness','var.Manice-thickness','var.CamdenA-thickness','var.CamdenB-thickness'),
           Thickness=c(6,7,5,5,6,5,7), # in mm
           end=c(11,13,13,12,9,14,11) # in mm
)
thickness<-ggplot(schambth,aes(x=Thickness,xend=end,y=Name,yend=Name,color=Name)) +
  geom_segment(size=2) +
  scale_color_brewer(palette = "Dark2") +
  theme(legend.position = "none")
schambstl<-data.frame(Name=c('var.Gary-stemlength','var.Malvern-stemlength','var.LeFlore-stemlength','var.Bodcaw-stemlength','var.Manice-stemlength','var.CamdenA-stemlength','var.CamdenB-stemlength'),
           StemLength=c(15,11,11,11,10,9,12), # in mm
           end=c(29,23,24,24,17,19,18) # in mm
)
stemlength<-ggplot(schambstl,aes(x=StemLength,xend=end,y=Name,yend=Name,color=Name)) +
  geom_segment(size=2) +
  scale_color_brewer(palette = "Dark2") +
  theme(legend.position = "none")
schambstw<-data.frame(Name=c('var.Gary-stemlength','var.Malvern-stemlength','var.LeFlore-stemlength','var.Bodcaw-stemlength','var.Manice-stemlength','var.CamdenA-stemlength'), # var.CamdenB-stemlength not listed in text
           StemWidth=c(20,17,13,15,12,11), # in mm
           end=c(28,25,33,24,24,21) # in mm
)
stemwidth<-ggplot(schambstw,aes(x=StemWidth,xend=end,y=Name,yend=Name,color=Name)) +
  geom_segment(size=2) +
  scale_color_brewer(palette = "Dark2") +
  theme(legend.position = "none")
# render figure
schambachfig<-ggarrange(length,width,thickness,stemlength,stemwidth,
                  labels = c("A","B","C","D","E"),
                  ncol = 2, nrow = 3)
annotate_figure(schambachfig,
                top=text_grob("Gary type-variety linear metrics per Schambach (1998)")
)
```

### Functions used to assign Gary type-varieties

```{r type function, echo=FALSE}
type1<-ppgary$frdwbgrp # to be deleted
# function to assign type-varieties based upon Ford and Webb (1956)
#ppgary$type1<-NA
#for (i in 1:(nrow(ppgary))){
#  ml <- maxl[i]
#
#  lcheck1 <- ml %in% 80:140
#  lcheck2 <- ml %in% 46:79
#  lcheck3 <- ml %in% 33:45
#
#  res <- if(lcheck1) "L" else if(lcheck2) "M/T" else
#    if(lcheck3) "S" else "UID"
#
#  ppgary$type1[i] <- res
#}
```

## Gary varieties proposed by Ford and Webb [-@RN20898]

The `type1` argument used within this and the subsequent section articulates with three variants of the Gary type (`Large`, `Typical`, and `Small`), which were assigned using those morphological criteria first advanced by Ford, Phillips, and Haag [-@RN20899] at the Jaketown site, and later refined by Ford and Webb [-@RN20898] at Poverty Point.

Each of the three type-varieties proposed by Ford and Webb [-@RN20898] was based upon a suite of morphological criteria that can be systematically replicated, and a function was used to assign each of the Gary points to the correct type-variety. Gary type varieties range between 80 and 140 mm in maxl, and between 30 and 42 mm in maxw for _Gary Large_; between 46 to 79 mm in maxl, and between 20 and 45 mm in maxw for _Gary Medium/Typical_; and between 33 to 45 mm in maxl, 19 to 32 mm in maxw, and five to 10 mm in maxth for _Gary Small_ [@RN20898].

### Boxplots for `variable` by `type1`

```{r boxplot, echo=TRUE}
# boxplot of maximum length ~ type
t1maxl<-ggplot(ppgary,aes(x=type1,y=maxl,color=type1)) + geom_boxplot() +
  geom_dotplot(binaxis = 'y',stackdir = 'center',dotsize = 0.3) +
  scale_color_brewer(palette = "Dark2") +
  theme(legend.position = "none")
# boxplot of maximum width ~ type
t1maxw<-ggplot(ppgary,aes(x=type1,y=maxw,color=type1)) + geom_boxplot() +
  geom_dotplot(binaxis = 'y',stackdir = 'center',dotsize = 0.3) +
  scale_color_brewer(palette = "Dark2") +
  theme(legend.position = "none")
# boxplot of maximum thickness ~ type
t1maxth<-ggplot(ppgary,aes(x=type1,y=maxth,color=type1)) + geom_boxplot() +
  geom_dotplot(binaxis = 'y',stackdir = 'center',dotsize = 0.3) +
  scale_color_brewer(palette = "Dark2") +
  theme(legend.position = "none")
# boxplot of stem length ~ type
t1maxstl<-ggplot(ppgary,aes(x=type1,y=maxstl,color=type1)) + geom_boxplot() +
  geom_dotplot(binaxis = 'y',stackdir = 'center',dotsize = 0.3) +
  scale_color_brewer(palette = "Dark2") +
  theme(legend.position = "none")
# boxplot of stem width ~ type
t1maxstw<-ggplot(ppgary,aes(x=type1,y=maxstw,color=type1)) + geom_boxplot() +
  geom_dotplot(binaxis = 'y',stackdir = 'center',dotsize = 0.3) +
  scale_color_brewer(palette = "Dark2") +
  theme(legend.position = "none")
# render figure
t1figure<-ggarrange(t1maxl,t1maxw,t1maxth,t1maxstl,t1maxstw,
                  labels = c("A","B","C","D","E"),
                  ncol = 3, nrow = 2)
t1figure
```

### Principal Components Analysis for `type1`

```{r pca, echo=TRUE}
#pca
ppgary.pca<-prcomp(ppgary[c(2:6)],center = TRUE,scale. = TRUE)
summary(ppgary.pca)
t1pca<-ggbiplot(ppgary.pca,obs.scale = 1,var.scale = 1,ellipse = TRUE,groups = type1) +
  scale_color_brewer(name = "Type-Variety",palette = "Dark2") +
  theme(legend.position = "right")
#render figure
t1pca
```

### Analyses of Variance (ANOVA) for `variable` ~ `type1`

```{r anova rrpp, echo=TRUE}
# anova = maximum length ~ type
t1ml<-lm.rrpp(maxl ~ type1, SS.type = "I",data = ppgary,iter = 9999,print.progress = FALSE)
anova(t1ml)
# anova = maximum width ~ type
t1mw<-lm.rrpp(maxw ~ type1, SS.type = "I",data = ppgary,iter = 9999,print.progress = FALSE)
anova(t1mw)
# anova = maximum thickness ~ type
t1mth<-lm.rrpp(maxth ~ type1, SS.type = "I",data = ppgary,iter = 9999,print.progress = FALSE)
anova(t1mth)
# anova = maximum stem length ~ type
t1mstl<-lm.rrpp(maxstl ~ type1, SS.type = "I",data = ppgary,iter = 9999,print.progress = FALSE)
anova(t1mstl)
# anova = maximum stem width ~ type
t1mstw<-lm.rrpp(maxw ~ type1, SS.type = "I",data = ppgary,iter = 9999,print.progress = FALSE)
anova(t1mstw)
# anova = dart-arrow index ~ type
t1dai<-lm.rrpp(dai ~ type1, SS.type = "I",data = ppgary,iter = 9999,print.progress = FALSE)
anova(t1dai)
```

## Acknowledgments

Funding for this project was provided to RZS by the United States Forest Service, National Forests and Grasslands in Texas (15-PA-11081300-033 and 20-PA-11081300-074), and components of this analytical work flow were developed and funded by a Preservation Technology and Training grant (P14AP00138) to RZS from the National Center for Preservation Technology and Training.

## References cited
